---
title: "Deconvolution with default methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deconvolution with default methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multideconv)
```

## **Deconvolution with default methods**

The basic function is to perform cell type deconvolution using six default methods (`quanTIseq`, `DeconRNASeq`, `CIBERSORTx`, `EpiDISH`, `DWLS`, `MOMF`) and nine default signatures (see paper [Hurtado et al., 2025](https://www.biorxiv.org/content/10.1101/2025.04.29.651220v2.article-info)). The function accepts either raw counts or TPM-normalized counts as input (with genes as SYMBOL).

**NOTE:** If you plan to use `CIBERSORTx`, you must provide your credentials (see README for details). The resulting deconvolution matrix is automatically saved in the `Results/` directory.

The output includes all combinations of deconvolution features, method-signature-cell type.

```{r, eval = FALSE}
bulk = multideconv::raw_counts
deconv = compute.deconvolution(raw.counts = bulk, 
                               methods = c("Quantiseq", "Epidish", 
                                           "DeconRNASeq", "DWLS","MOMF"), 
                               normalized = TRUE, 
                               return = TRUE, 
                               file_name = "Tutorial")
```

To run only specific signatures, use the `signatures_select` argument. If `NULL` (default), all available signatures (package + custom from `Results/custom_signatures/`) are used:

```{r, eval = FALSE}
deconv = compute.deconvolution(raw.counts = bulk,
                               methods = c("Quantiseq", "DeconRNASeq"),
                               normalized = TRUE,
                               signatures_select = c("BPRNACan", "LM22"),
                               return = TRUE,
                               file_name = "Tutorial")
```

To speed up computation, `multideconv` supports parallelization. Set `doParallel = TRUE` and specify the number of workers based on your system's resources:

```{r, eval = FALSE}
deconv = compute.deconvolution(raw.counts = bulk, 
                               methods = "DWLS", 
                               normalized = TRUE, 
                               return = TRUE, 
                               file_name = "Tutorial", 
                               doParallel = TRUE, 
                               workers = 3)
```

Once deconvolution is complete, pass the result to `compute.deconvolution.analysis()` to reduce redundancy, group correlated features into subgroups, and (optionally) correct for batch/cohort effects — see the [Cell type subgroup analysis](a4_subgroups.html) article for a full walkthrough including the `batch` parameter and pathway characterisation with `compute.subgroup.pathways()`.

## **Cell type signatures**

In order to access the default signatures `multideconv` provides, you can do the following:

To list all signatures

```{r}
path <- system.file("signatures/", package = "multideconv")
list.files(path)
```

To access a specific signature

```{r}
signature = read.delim(paste0(path, "CBSX-Melanoma-scRNAseq.txt"))
head(signature)
```
