---
title: "Single-cell deconvolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-cell deconvolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multideconv)
```

## **Single-cell metacell construction**

If single-cell data is available, we recommend generating metacells to reduce computation time and prevent session crashes. Deconvolution methods that rely on single-cell data can be computationally intensive, especially with large matrices. We suggest using a maximum of 20k cells; if your object exceeds this size, creating metacells is strongly advised. However, if your computational resources are sufficient to handle the full single-cell dataset, you may skip this step.

We adapted functions from the R package hdWGCNA (@morabito2023hdwgcna; @langfelder2008wgcna) for the construction of metacells using the KNN algorithm.

- **sc_object**: Normalized gene expression matrix with genes as rows and cells as columns
- **labels_column**: Vector of cell annotations
- **samples_column**: Vector of sample IDs for each cell
- **exclude_cells**: Vector specifying which cell types to ignore during metacell construction (default is NULL)
- **min_cells**: Minimum number of cells required to construct metacells in a group
- **k**: Number of nearest neighbors used for the KNN algorithm
- **max_shared**: Maximum number of cells shared between two metacells
- **n_workers**: Number of cores to use for parallelizing metacell construction
- **min_meta**: Minimum number of metacells required for a cell type to be retained

Because of space limitations, we have not included a complete single-cell object in this tutorial. However, users are expected to provide their own single-cell data and supply it to the `sc_object` parameter in the function call.

**NOTE:** `create_metacells()` requires the `hdWGCNA` package, which is not installed automatically. Install it once before first use:

```{r, eval=FALSE}
# install.packages("pak")
pak::pkg_install("smorabit/hdWGCNA")
```

```{r, eval=FALSE}
metacells = create_metacells(sc_object,
                             labels_column = cell_labels, 
                             samples_column = sample_labels, 
                             exclude_cells = NULL,
                             min_cells = 50, 
                             k = 15, 
                             max_shared = 15, 
                             n_workers = 4, 
                             min_meta = 10)
```

## **Second-generation deconvolution methods**

Once the single-cell data is prepared, users can supplement the default deconvolution methods with second-generation approaches such as `AutogeneS`, `BayesPrism`, `Bisque`, `CPM`, `MuSic`, and `SCDC`. These methods learn cell-type signatures directly from annotated single-cell RNA-seq data, rather than relying on predefined static signatures (@Dietrich2024.06.10.598226), to deconvolve bulk RNA-seq profiles.

- **sc_deconv**: Boolean indicating whether to run second-generation methods
- **sc_matrix**: Normalized single-cell gene expression matrix
- **sc_metadata**: Dataframe containing single-cell metadata
- **cell_label**: Column name in `sc_metadata` with cell type annotations
- **sample_label**: Column name in `sc_metadata` with sample IDs
- **name_sc_signature**: Name to assign to the resulting signature

```{r}
metacell_obj = multideconv::metacells_data
metacell_metadata = multideconv::metacells_metadata
head(metacell_obj[1:5,1:5])
head(metacell_metadata)
```

This function computes cell type deconvolution using the six default methods (`quanTIseq`, `DeconRNASeq`, `EpiDISH`, `DWLS`, `MOMF`) and `CIBERSORTx` (if credentials are provided), along with second-generation deconvolution approaches. The output includes all combinations of methods and signatures.

```{r, eval=FALSE}
deconv = compute.deconvolution(raw.counts = bulk, 
                               normalized = TRUE, 
                               return = TRUE, 
                               methods = c("Quantiseq", "Epidish", "DeconRNASeq"),
                               file_name = "Tutorial", 
                               sc_deconv = TRUE, 
                               sc_matrix = metacell_obj, 
                               sc_metadata = metacell_metadata, 
                               methods_sc = c("Autogenes", "BayesPrism", 
                                              "Bisque", "CPM", "MuSic", "SCDC"),
                               cell_label = "annotated_ct", 
                               sample_label = "sample", 
                               name_sc_signature = "Test")
```

To run only the second-generation deconvolution methods based on single-cell data, without using any static cell-type signatures, use the public `compute.deconvolution()` workflow with an empty first-generation `methods` vector:

```{r, eval=FALSE}
deconv_sc = compute.deconvolution(raw.counts = bulk,
                                  normalized = TRUE,
                                  methods = character(0),
                                  sc_deconv = TRUE,
                                  sc_matrix = metacell_obj,
                                  sc_metadata = metacell_metadata,
                                  methods_sc = c("Autogenes", "BayesPrism",
                                                 "Bisque", "CPM", "MuSic", "SCDC"),
                                  cell_label = "annotated_ct",
                                  sample_label = "sample",
                                  name_sc_signature = "Test",
                                  workers = 2,
                                  return = TRUE,
                                  file_name = "Tutorial")
```
