---
title: "Cell type subgroup analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell type subgroup analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multideconv)
```

## **Cell type processing**

Deconvolution analysis reduces the dimensionality and heterogeneity of the deconvolution results. It uses the cell type processing algorithm described in the paper [Hurtado et al., 2025](https://www.biorxiv.org/content/10.1101/2025.04.29.651220v2.article-info). It returns the cell type subgroups composition and the reduced deconvolution matrix, saved in the `Results/` directory.

-   **deconvolution**: Matrix of raw deconvolution results (output of `compute.deconvolution()`)

-   **corr**: Minimum correlation threshold to group features

-   **seed**: Random seed for reproducibility

-   **return**: Whether to return results and save output files to the `Results/` directory

```{r, message=FALSE}
deconv_bulk = multideconv::deconv_bulk
deconv_subgroups = compute.deconvolution.analysis(deconvolution = deconv_bulk, 
                                                  corr = 0.7, 
                                                  seed = 123, 
                                                  file_name = "Tutorial", 
                                                  return = TRUE) 
```

Subgroups composition can be extracted with:

```{r}
deconv_subgroups[[3]]$B.cells
deconv_subgroups[[3]]$Macrophages.M2
deconv_subgroups[[3]]$Dendritic.cells
```

Reduced deconvolution matrix:

```{r}
head(subgroups[[1]][,sample(colnames(subgroups[[1]]), 10)])
```

If your deconvolution matrix contains non-standard cell types (see README), specify them using `cells_extra` to ensure proper subgrouping. If not, they are going to be discarded automatically.

```{r, eval = FALSE}
deconv_subgroups = compute.deconvolution.analysis(deconvolution = deconv_pseudo, 
                                                  corr = 0.7, 
                                                  seed = 123, 
                                                  return = TRUE,
                                                  cells_extra = c("Mural.cells", "Myeloid.cells"), 
                                                  file_name = "Tutorial") 
```

## **Deconvolution dictionary**

The deconvolution dictionary step integrates deconvolution features with pathway activity information to provide a functional interpretation of each cell-type–specific component. This process identifies groups of pathways that show coordinated behavior across samples (e.g., immunoactive vs. immunosuppressive signaling) and assigns each deconvolution feature to one of these pathway clusters based on correlation patterns. The resulting dictionary can then be used to understand whether a cell-type feature is more associated with activation or suppression programs.

For this, we start from normalized gene expression data and link it to deconvolution results obtained in previous steps. Finally, we use the `compute_deconvolution_dictionary()` function to classify each deconvolution feature according to its dominant pathway program.

```{r, message=FALSE, warning=FALSE}
counts = multideconv::raw_counts
counts.norm = ADImpute::NormalizeTPM(counts, log = FALSE)
deconv = multideconv::deconv_bulk[rownames(multideconv::deconv_bulk)%in%colnames(counts),]
deconv_subgroups = compute.deconvolution.analysis(deconvolution = deconv, 
                                                  corr = 0.7, 
                                                  seed = 123, 
                                                  file_name = "Tutorial", 
                                                  return = TRUE) 
deconv_dictionary = compute_deconvolution_dictionary(subgroups = deconv_subgroups,
                                                     expr = counts.norm)
```

The dictionary matrix now contains pathway-annotated features:

```{r}
head(deconv_dictionary[["Deconvolution matrix"]][,1:5])
```

The dictionary also identifies global pathway states (immunoactive, immunosuppressive, mixed):

```{r}
deconv_dictionary[["States"]]
```

## **Pathway relationship visualization**

To visualize pathway relationships for the learned subgroup structure, use `compute_subgroups_pathways()`. This function computes PROGENy pathway activity from the normalized counts and generates module relationship plots showing how well each deconvolution subgroup correlates with each pathway activity pattern.

```{r, message=FALSE, warning=FALSE, eval = FALSE}
# Generate pathway relationship plots for all cell types
# Plots will be saved to Results/ directory
compute_subgroups_pathways(subgroups = deconv_dictionary,
                           counts_norm = counts.norm,
                           file_name = "Tutorial",
                           height = 6,
                           width = 12,
                           pval = 0.05)
```

The function generates one PDF plot per cell type. Each plot shows the module relationship matrix between deconvolution subgroups and PROGENy pathway activity patterns, with correlation strength and statistical significance displayed. 

**Example plots generated for selected cell types:**

The plots for B cells and Macrophages are saved as:
- `Results/Tutorial_B.cells.pdf` 
- `Results/Tutorial_Macrophages.M0.pdf`

These visualizations help interpret which pathway programs are co-active with each cell type's deconvolution subgroups, revealing the functional annotation of the identified subgroups.

## **Replicate deconvolution subgroups in an independent set**

Cell subgroup identification through deconvolution is cohort-specific, as it relies on correlation patterns across samples. This means that subgroup definitions may vary across different splits or datasets. If you aim to replicate the same subgroups identified in one dataset onto another (e.g., for model validation), you can use the following function.

The function below reconstructs and applies the subgroup signatures derived from a previous deconvolution, making it especially useful when transferring learned patterns across datasets—such as when training and evaluating machine learning models. 
```{r}
deconv_1 = deconv_bulk[1:100,]
deconv_2 = deconv_bulk[101:192,]

deconv_subgroups = compute.deconvolution.analysis(deconvolution = deconv_1, 
                                                  corr = 0.7, 
                                                  seed = 123, 
                                                  file_name = "Tutorial", 
                                                  return = FALSE) 
deconv_subgroups_replicate = replicate_deconvolution_subgroups(deconv_subgroups, 
                                                               deconv_2)
```
