---
title: "Machine learning workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine learning workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## **Machine learning models using deconvolution subgroups**

The deconvolution subgroups generated by `multideconv` can be used as input features for training machine learning (ML) models. However, since these subgroups are derived based on sample-level correlations, special care is needed to avoid data leakage. If you compute the full subgroup matrix on the entire dataset before splitting into training and test sets (or before performing k-fold cross-validation), the model may indirectly access information from the test set during training — a form of hidden data leakage described in @hurtado2026pipeline.

To address this issue, we provide the function `prepare_multideconv_folds()`, which ensures proper separation between training and test data during the computation of deconvolution subgroups. Within each cross-validation fold, the subgroups are learned only from the training samples of the fold and then projected onto its test samples with `replicate_deconvolution_subgroups()`.

Below is an example using `pipeML`, an R package that supports custom fold construction: `prepare_multideconv_folds()` is passed to the argument `fold_construction_fun`. For more information about the function arguments, visit the documentation of [pipeML](https://github.com/VeraPancaldiLab/pipeML).

### Input data

- `deconv_train`: deconvolution of the training samples (samples as rows, deconvolution features as columns), as returned by `compute.deconvolution()`.
- `deconv_test`: deconvolution of the test samples, with the same features.
- `traitData_train`, `traitData_test`: clinical data of the training and test samples, with the response in a `Response` column (`"R"` for responders). Rows must be in the same order as the deconvolution tables.

```{r, eval = FALSE}
library(pipeML)
library(multideconv)

# Make sure the clinical data follows the order of the deconvolution tables
traitData_train <- traitData_train[rownames(deconv_train), , drop = FALSE]
traitData_test  <- traitData_test[rownames(deconv_test), , drop = FALSE]
```

### Training

`compute_features.training.ML()` builds the cross-validation folds and calls `prepare_multideconv_folds()` to compute the deconvolution subgroups of each fold. Arguments of `prepare_multideconv_folds()` (e.g. `ncores`, `corr`, `corr_type`, `zero_thr`, `var_quantile`, `prune_thr`) can be set with `fold_construction_args_fixed`; they are applied in every fold and in the final model:

```{r, eval = FALSE}
res <- pipeML::compute_features.training.ML(features_train = deconv_train,
                                            target_var = traitData_train$Response,
                                            task_type = "classification",
                                            trait.positive = "R",
                                            metric = "AUROC",
                                            k_folds = 5,
                                            n_rep = 10,
                                            ncores = 3,
                                            return = FALSE,
                                            fold_construction_fun = prepare_multideconv_folds,
                                            fold_construction_args_fixed = list(ncores = 3))
```

- `res$Model` is the selected model, trained on the deconvolution subgroups computed on all the training samples.
- `res$Custom_output` is the output of `compute.deconvolution.analysis()` on all the training samples: the subgroups used by the final model.

### Prediction

Project the test samples onto the deconvolution subgroups learned on the training samples, then predict:

```{r, eval = FALSE}
# Replicate deconvolution subgroups in the test samples
dt_test <- multideconv::replicate_deconvolution_subgroups(deconv_res = res$Custom_output,
                                                          deconvolution_test = deconv_test)

# Predict on the test samples
pred <- pipeML::compute_prediction(model = res$Model,
                                   test_data = dt_test,
                                   target_var = traitData_test$Response,
                                   task_type = "classification",
                                   trait.positive = "R",
                                   return = TRUE)

pred$AUC   # AUROC and AUPRC with bootstrap confidence intervals
```

With `return = TRUE`, ROC and PR curves are saved in the `Results/` folder.

### Which deconvolution subgroups drive the predictions?

`compute_shap_values()` computes SHAP values for the final model (`res$Model`), trained on the deconvolution subgroups computed on all the training samples (`res$Custom_output`). Each subgroup therefore has the same definition for all samples. Everything else is taken from the trained model:

```{r, eval = FALSE}
shap <- pipeML::compute_shap_values(model_trained = res$Model,
                                    task_type = "classification")

# Global importance of each deconvolution subgroup (mean |SHAP| across samples)
head(sort(colMeans(abs(shap)), decreasing = TRUE), 10)
```

SHAP values are in units of predicted probability of response: a positive value means the subgroup pushed the prediction of that sample towards response. The final model was trained on the samples it explains, so SHAP values describe how the model uses each subgroup on the training data; compare its training performance with the cross-validation performance to check that it does not overfit. See the `pipeML` vignette for plots of SHAP values per sample and across samples.

### Survival outcomes

`prepare_multideconv_folds()` also works for survival analysis. `pipeML` gives it the survival time and event of the training samples (in the `time` and `event` columns of its `data` argument), so no extra argument is needed:

```{r, eval = FALSE}
res_survival <- pipeML::compute_features.training.ML(features_train = deconv_train,
                                                     task_type = "survival",
                                                     time_var = traitData_train$time,
                                                     event_var = traitData_train$event,   # 1 = event, 0 = censored
                                                     k_folds = 5,
                                                     n_rep = 10,
                                                     ncores = 3,
                                                     fold_construction_fun = prepare_multideconv_folds,
                                                     fold_construction_args_fixed = list(ncores = 3))

dt_test <- multideconv::replicate_deconvolution_subgroups(deconv_res = res_survival$Custom_output,
                                                          deconvolution_test = deconv_test)

pred_survival <- pipeML::compute_prediction(model = res_survival$Model,
                                            test_data = dt_test,
                                            task_type = "survival",
                                            time_var = traitData_test$time,
                                            event_var = traitData_test$event)
pred_survival$c_index
```

**NOTE:** `multideconv` is built on top of existing frameworks and makes extensive use of the R packages `immunedeconv` (@10.1093/bioinformatics/btz363) and `omnideconv` (@Dietrich2024.06.10.598226). If you use `multideconv` in your work, please cite our package along with these foundational packages. We also encourage you to cite the individual deconvolution algorithms you employ in your analysis.

| method | license | citation |
|-----------------|-----------------|---------------------------------------|
| [quanTIseq](http://icbi.at/software/quantiseq/doc/index.html) | free ([BSD](https://github.com/omnideconv/immunedeconv/blob/master/LICENSE.md)) | Finotello, F., Mayer, C., Plattner, C., Laschober, G., Rieder, D., Hackl, H., ..., Sopper, S. (2019). Molecular and pharmacological modulators of the tumor immune contexture revealed by deconvolution of RNA-seq data. Genome medicine, 11(1), 34. <https://doi.org/10.1186/s13073-019-0638-6> |
| [EpiDISH](https://github.com/sjczheng/EpiDISH) | free ([GPL 2.0](https://github.com/sjczheng/EpiDISH/blob/devel/DESCRIPTION)) | Zheng SC, Breeze CE, Beck S, Teschendorff AE (2018). "Identification of differentially methylated cell-types in Epigenome-Wide Association Studies." Nature Methods, 15(12), 1059. <https://doi.org/10.1038/s41592-018-0213-x> |
| [DeconRNASeq](https://git.bioconductor.org/packages/DeconRNASeq) | free ([GPL-2](https://www.bioconductor.org/packages/release/bioc/html/DeconRNASeq.html)) | joseph.szustakowski@novartis.com TGJDS (2025). DeconRNASeq: Deconvolution of Heterogeneous Tissue Samples for mRNA-Seq data. <doi:10.18129/B9.bioc.DeconRNASeq>, R package version 1.50.0, <https://bioconductor.org/packages/DeconRNASeq> |
| [AutoGeneS](https://github.com/theislab/AutoGeneS/) | free ([MIT](https://github.com/theislab/AutoGeneS/blob/master/LICENSE)) | Aliee, H., & Theis, F. (2021). AutoGeneS: Automatic gene selection using multi-objective optimization for RNA-seq deconvolution. <https://doi.org/10.1101/2020.02.21.940650> |
| [BayesPrism](https://github.com/Danko-Lab/BayesPrism) | free ([GPL 3.0](https://github.com/Danko-Lab/BayesPrism/blob/main/BayesPrism/DESCRIPTION)) | Chu, T., Wang, Z., Pe'er, D. et al. Cell type and gene expression deconvolution with BayesPrism enables Bayesian integrative analysis across bulk and single-cell RNA sequencing in oncology. Nat Cancer 3, 505–517 (2022). <https://doi.org/10.1038/s43018-022-00356-3> |
| [Bisque](https://github.com/cozygene/bisque) | free ([GPL 3.0](https://github.com/cozygene/bisque/blob/master/DESCRIPTION)) | Jew, B., Alvarez, M., Rahmani, E., Miao, Z., Ko, A., Garske, K. M., Sul, J. H., Pietiläinen, K. H., Pajukanta, P., & Halperin, E. (2020). Publisher Correction: Accurate estimation of cell composition in bulk expression through robust integration of single-cell information. Nature Communications, 11(1), 2891. <https://doi.org/10.1038/s41467-020-16607-9> |
| [BSeq-sc](https://github.com/shenorrLab/bseqsc) | free ([GPL 2.0](https://github.com/shenorrLab/bseqsc/blob/master/DESCRIPTION)) | Baron, M., Veres, A., Wolock, S. L., Faust, A. L., Gaujoux, R., Vetere, A., Ryu, J. H., Wagner, B. K., Shen-Orr, S. S., Klein, A. M., Melton, D. A., & Yanai, I. (2016). A Single-Cell Transcriptomic Map of the Human and Mouse Pancreas Reveals Inter- and Intra-cell Population Structure. In Cell Systems (Vol. 3, Issue 4, pp. 346–360.e4). <https://doi.org/10.1016/j.cels.2016.08.011> |
| [CIBERSORTx](https://cibersortx.stanford.edu/) | free for non-commerical use only | Newman, A. M., Liu, C. L., Green, M. R., Gentles, A. J., Feng, W., Xu, Y., Hoang, C. D., Diehn, M., & Alizadeh, A. A. (2015). Robust enumeration of cell subsets from tissue expression profiles. Nature Methods, 12(5), 453–457. <https://doi.org/10.1038/s41587-019-0114-2> |
| [CPM](https://github.com/amitfrish/scBio) | free ([GPL 2.0](https://github.com/amitfrish/scBio/blob/master/DESCRIPTION)) | Frishberg, A., Peshes-Yaloz, N., Cohn, O., Rosentul, D., Steuerman, Y., Valadarsky, L., Yankovitz, G., Mandelboim, M., Iraqi, F. A., Amit, I., Mayo, L., Bacharach, E., & Gat-Viks, I. (2019). Cell composition analysis of bulk genomics using single-cell data. Nature Methods, 16(4), 327–332. <https://doi.org/10.1038/s41592-019-0355-5> |
| [DWLS](https://bitbucket.org/yuanlab/dwls/src/master/) | free ([GPL](https://bitbucket.org/yuanlab/dwls/src/master/DESCRIPTION)) | Tsoucas, D., Dong, R., Chen, H., Zhu, Q., Guo, G., & Yuan, G.-C. (2019). Accurate estimation of cell-type composition from gene expression data. Nature Communications, 10(1), 2975. <https://doi.org/10.1038/s41467-019-10802-z> |
| [MOMF](https://github.com/sqsun/MOMF) | free ([GPL 3.0](https://github.com/sqsun/MOMF/blob/master/LICENSE.md)) | Xifang Sun, Shiquan Sun, and Sheng Yang. An efficient and flexible method for deconvoluting bulk RNAseq data with single-cell RNAseq data, 2019, DOI: 10.5281/zenodo.3373980 |
| [MuSiC](https://github.com/xuranw/MuSiC/) | free ([GPL 3.0](https://github.com/xuranw/MuSiC/blob/master/LICENSE)) | Wang, X., Park, J., Susztak, K., Zhang, N. R., & Li, M. (2019). Bulk tissue cell type deconvolution with multi-subject single-cell expression reference. Nature Communications, 10(1), 380. <https://doi.org/10.1038/s41467-018-08023-x> |
| [SCDC](https://github.com/meichendong/SCDC) | ([MIT](https://github.com/meichendong/SCDC/blob/master/README.md)) | Dong, M., Thennavan, A., Urrutia, E., Li, Y., Perou, C. M., Zou, F., & Jiang, Y. (2020). SCDC: bulk gene expression deconvolution by multiple single-cell RNA sequencing references. Briefings in Bioinformatics. <https://doi.org/10.1093/bib/bbz166> |

# References
