---
title: "multideconv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multideconv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette presents a practical, end-to-end workflow for cell type deconvolution with `multideconv`, including first-generation and second-generation methods, subgroup processing, benchmarking, and downstream machine learning use.

## Tutorial {.tabset .tabset-pills}

### Get started

This tab contains README-level onboarding for installation and basic usage.

For the complete practical workflow, open the **Articles** tab.

#### Installation

To avoid GitHub API rate limit issues during installation, configure a GitHub Personal Access Token (PAT):

```{r}
# install.packages(c("usethis", "gitcreds"))
usethis::create_github_token()
gitcreds::gitcreds_set()
```

Install `multideconv` from GitHub:

```{r}
# install.packages("pak")
pak::pkg_install("VeraPancaldiLab/multideconv")
```

#### Basic usage

All outputs are written to a local `Results/` folder in your working directory.

```{r}
setwd("~/path/to/project")
library(multideconv)
raw.counts <- multideconv::raw_counts

deconv <- compute.deconvolution(
  raw.counts,
  normalized = TRUE,
  file_name = "Tutorial"
)
```

`CIBERSORTx` requires valid credentials and Docker configured for non-root use.

### Articles

#### 1) Run deconvolution with default signatures

`compute.deconvolution()` combines multiple methods and signatures on bulk RNA-seq expression data (raw counts or TPM-normalized values with gene symbols).

```{r}
bulk <- multideconv::raw_counts

deconv <- compute.deconvolution(
  raw.counts = bulk,
  methods = c("Quantiseq", "Epidish", "DeconRNASeq", "DWLS", "MOMF"),
  normalized = TRUE,
  return = TRUE,
  file_name = "Tutorial"
)
```

Use `methods`, `methods_sc`, `signatures_exclude`, and `doParallel` to customize runtime and scope.

```{r}
deconv <- compute.deconvolution(
  raw.counts = bulk,
  methods = c("Quantiseq", "DeconRNASeq"),
  signatures_exclude = "BPRNACan",
  normalized = TRUE,
  return = TRUE,
  file_name = "Tutorial"
)
```

```{r}
deconv <- compute.deconvolution(
  raw.counts = bulk,
  methods = "DWLS",
  normalized = TRUE,
  doParallel = TRUE,
  workers = 3,
  return = TRUE,
  file_name = "Tutorial"
)
```

#### 2) Inspect available signatures

```{r}
path <- system.file("signatures/", package = "multideconv")
list.files(path)

signature <- read.delim(paste0(path, "CBSX-Melanoma-scRNAseq.txt"))
head(signature)
```

#### 3) Prepare single-cell inputs and metacells

For large scRNA-seq references, metacells can reduce memory use and execution time before running second-generation methods.

```{r}
metacells <- create_metacells(
  sc_object,
  labels_column = cell_labels,
  samples_column = sample_labels,
  exclude_cells = NULL,
  min_cells = 50,
  k = 15,
  max_shared = 15,
  n_workers = 4,
  min_meta = 10
)
```

Example built-in data:

```{r}
metacell_obj <- multideconv::metacells_data
metacell_metadata <- multideconv::metacells_metadata
head(metacell_obj[1:5, 1:5])
head(metacell_metadata)
```

#### 4) Run second-generation deconvolution

```{r}
deconv <- compute.deconvolution(
  raw.counts = bulk,
  normalized = TRUE,
  methods = c("Quantiseq", "Epidish", "DeconRNASeq"),
  sc_deconv = TRUE,
  sc_matrix = metacell_obj,
  sc_metadata = metacell_metadata,
  methods_sc = c("Autogenes", "BayesPrism", "Bisque", "CPM", "MuSic", "SCDC"),
  cell_label = "annotated_ct",
  sample_label = "sample",
  name_sc_signature = "Test",
  return = TRUE,
  file_name = "Tutorial"
)
```

To compute only second-generation methods:

```{r}
deconv_sc <- compute_sc_deconvolution_methods(
  raw_counts = bulk,
  normalized = TRUE,
  methods_sc = c("Autogenes", "BayesPrism", "Bisque", "CPM", "MuSic", "SCDC"),
  sc_object = metacell_obj,
  sc_metadata = metacell_metadata,
  cell_annotations = "annotated_ct",
  samples_ids = "sample",
  name_object = "Test",
  n_cores = 2,
  return = TRUE,
  file_name = "Tutorial"
)
```

#### 5) Build pseudo-bulk and benchmark signatures

```{r}
metacells_seurat <- Seurat::CreateSeuratObject(metacell_obj, meta.data = metacell_metadata)
pseudobulk <- create_sc_pseudobulk(
  metacells_seurat,
  cells_labels = "annotated_ct",
  sample_labels = "sample",
  normalized = TRUE,
  file_name = "Tutorial"
)
```

Generate signatures from scRNA-seq:

```{r}
signatures <- create_sc_signatures(
  metacell_obj,
  metacell_metadata,
  cells_labels = "annotated_ct",
  sample_labels = "sample",
  bulk_rna = multideconv::pseudobulk,
  cell_markers = NULL,
  name_signature = "Test",
  methods_sig = c("DWLS", "CIBERSORTx", "MOMF", "BSeqsc")
)
```

Benchmark against known cell proportions:

```{r}
benchmark <- compute.benchmark(
  multideconv::deconvolution,
  multideconv::cells_groundtruth,
  cells_extra = c("Mural.cells", "Myeloid.cells"),
  corr_type = "spearman",
  scatter = FALSE,
  plot = TRUE,
  pval = 0.05,
  file_name = "Tutorial",
  width = 10,
  height = 15
)
```

#### 6) Process cell types and derive subgroup features

`compute.deconvolution.analysis()` reduces redundancy across deconvolution outputs and builds subgroup-level representations for downstream analysis.

```{r}
deconv_bulk <- multideconv::deconv_bulk

deconv_subgroups <- compute.deconvolution.analysis(
  deconvolution = deconv_bulk,
  corr = 0.7,
  seed = 123,
  return = TRUE,
  file_name = "Tutorial"
)
```

Include custom cell names when needed:

```{r}
deconv_subgroups <- compute.deconvolution.analysis(
  deconvolution = multideconv::deconvolution,
  corr = 0.7,
  seed = 123,
  cells_extra = c("Mural.cells", "Myeloid.cells"),
  return = TRUE,
  file_name = "Tutorial"
)
```

#### 7) Create a deconvolution dictionary

This step links deconvolution-derived features with pathway activity states.

```{r}
counts <- multideconv::raw_counts
counts.norm <- ADImpute::NormalizeTPM(counts, log = TRUE)
pathways <- CellTFusion::compute.pathway.activity(counts.norm)

deconv <- multideconv::deconv_bulk[rownames(multideconv::deconv_bulk) %in% colnames(counts), ]
deconv_subgroups <- compute.deconvolution.analysis(
  deconvolution = deconv,
  corr = 0.7,
  seed = 123,
  return = TRUE,
  file_name = "Tutorial"
)

deconv_subgroups <- deconvolution_dictionary(deconv_subgroups, pathways)
head(deconv_subgroups[[1]][, 1:5])
print(deconv_subgroups[["States"]])
```

#### 8) Replicate subgroup definitions in an external cohort

```{r}
deconv_1 <- deconv_bulk[1:100, ]
deconv_2 <- deconv_bulk[101:192, ]

subgroups_train <- compute.deconvolution.analysis(
  deconvolution = deconv_1,
  corr = 0.7,
  seed = 123,
  return = FALSE,
  file_name = "Tutorial"
)

deconv_subgroups_replicate <- replicate_deconvolution_subgroups(subgroups_train, deconv_2)
```

#### 9) Use subgroup features in machine learning workflows

To avoid data leakage, build subgroup features inside each training fold and then project to test folds.

```{r}
library(pipeML)

# traitData_train / traitData_test: clinical labels
# deconv_train / deconv_test: deconvolution matrices
res <- pipeML::compute_features.training.ML(
  deconv_train,
  traitData_train$Response,
  trait.positive = "R",
  metric = "AUROC",
  stack = FALSE,
  k_folds = 5,
  n_rep = 10,
  feature.selection = FALSE,
  LODO = FALSE,
  ncores = 3,
  return = FALSE,
  fold_construction_fun = prepare_multideconv_folds
)

dt_test <- replicate_deconvolution_subgroups(
  res$Custom_output$Processed_deconvolution,
  deconv_test
)

pred <- pipeML::compute_prediction(
  res,
  dt_test,
  traitData_test$Response,
  trait.positive = "R",
  stack = FALSE,
  maximize = "Accuracy",
  return = TRUE
)
```

#### Practical notes

- If you plan to use `CIBERSORTx`, provide valid credentials through `credentials.mail` and `credentials.token`.
- Ensure Docker is installed and available for non-root execution if `CIBERSORTx` is enabled.
- For patched compatibility with `omnideconv`, use:

```{r}
devtools::install_github("omnideconv/MuSiC")
devtools::install_github("omnideconv/SCDC")
```

- If you use `multideconv` in publications, cite `multideconv` and the underlying deconvolution frameworks and methods used in your analysis.

### References

#### Core method list

| method | license | citation |
|-----------------|-----------------|---------------------------------------|
| [quanTIseq](http://icbi.at/software/quantiseq/doc/index.html) | free ([BSD](https://github.com/omnideconv/immunedeconv/blob/master/LICENSE.md)) | Finotello, F., Mayer, C., Plattner, C., Laschober, G., Rieder, D., Hackl, H., ..., Sopper, S. (2019). Molecular and pharmacological modulators of the tumor immune contexture revealed by deconvolution of RNA-seq data. Genome medicine, 11(1), 34. <https://doi.org/10.1186/s13073-019-0638-6> |
| [EpiDISH](https://github.com/sjczheng/EpiDISH) | free ([GPL 2.0](https://github.com/sjczheng/EpiDISH/blob/devel/DESCRIPTION)) | Zheng SC, Breeze CE, Beck S, Teschendorff AE (2018). “Identification of differentially methylated cell-types in Epigenome-Wide Association Studies.” Nature Methods, 15(12), 1059. <https://doi.org/10.1038/s41592-018-0213-x> |
| [DeconRNASeq](https://git.bioconductor.org/packages/DeconRNASeq) | free ([GPL-2](https://www.bioconductor.org/packages/release/bioc/html/DeconRNASeq.html)) | Szustakowski, J. et al. DeconRNASeq: Deconvolution of Heterogeneous Tissue Samples for mRNA-Seq data. Bioconductor package. |
| [AutoGeneS](https://github.com/theislab/AutoGeneS/) | free ([MIT](https://github.com/theislab/AutoGeneS/blob/master/LICENSE)) | Aliee, H., & Theis, F. (2021). AutoGeneS: Automatic gene selection using multi-objective optimization for RNA-seq deconvolution. <https://doi.org/10.1101/2020.02.21.940650> |
| [BayesPrism](https://github.com/Danko-Lab/BayesPrism) | free ([GPL 3.0](https://github.com/Danko-Lab/BayesPrism/blob/main/BayesPrism/DESCRIPTION)) | Chu, T., Wang, Z., Pe’er, D. et al. Nat Cancer 3, 505–517 (2022). <https://doi.org/10.1038/s43018-022-00356-3> |
| [Bisque](https://github.com/cozygene/bisque) | free ([GPL 3.0](https://github.com/cozygene/bisque/blob/master/DESCRIPTION)) | Jew, B. et al. Nature Communications, 11(1), 2891 (2020). <https://doi.org/10.1038/s41467-020-16607-9> |
| [BSeq-sc](https://github.com/shenorrLab/bseqsc) | free ([GPL 2.0](https://github.com/shenorrLab/bseqsc/blob/master/DESCRIPTION)) | Baron, M. et al. Cell Systems, 3(4), 346–360.e4 (2016). <https://doi.org/10.1016/j.cels.2016.08.011> |
| [CIBERSORTx](https://cibersortx.stanford.edu/) | free for non-commercial use | Newman, A. M. et al. Nature Methods, 12(5), 453–457 (2015). |
| [CPM](https://github.com/amitfrish/scBio) | free ([GPL 2.0](https://github.com/amitfrish/scBio/blob/master/DESCRIPTION)) | Frishberg, A. et al. Nature Methods, 16(4), 327–332 (2019). <https://doi.org/10.1038/s41592-019-0355-5> |
| [DWLS](https://bitbucket.org/yuanlab/dwls/src/master/) | free ([GPL](https://bitbucket.org/yuanlab/dwls/src/master/DESCRIPTION)) | Tsoucas, D. et al. Nature Communications, 10(1), 2975 (2019). <https://doi.org/10.1038/s41467-019-10802-z> |
| [MOMF](https://github.com/sqsun/MOMF) | free ([GPL 3.0](https://github.com/sqsun/MOMF/blob/master/LICENSE.md)) | Sun, X., Sun, S., & Yang, S. (2019). <https://doi.org/10.5281/zenodo.3373980> |
| [MuSiC](https://github.com/xuranw/MuSiC/) | free ([GPL 3.0](https://github.com/xuranw/MuSiC/blob/master/LICENSE)) | Wang, X. et al. Nature Communications, 10(1), 380 (2019). <https://doi.org/10.1038/s41467-018-08023-x> |
| [SCDC](https://github.com/meichendong/SCDC) | free ([MIT](https://github.com/meichendong/SCDC/blob/master/README.md)) | Dong, M. et al. Briefings in Bioinformatics (2020). <https://doi.org/10.1093/bib/bbz166> |

# References
