---
title: "multideconv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multideconv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`multideconv` is an integrative pipeline for combining first- and second-generation cell type deconvolution methods from bulk RNA-seq data, as described in [Hurtado et al., 2025](https://www.biorxiv.org/content/10.1101/2025.04.29.651220v2.article-info).

## Installation

To avoid GitHub API rate limit issues, set up a Personal Access Token (PAT) before installing:

```{r, eval=FALSE}
# install.packages(c("usethis", "gitcreds"))
usethis::create_github_token()
gitcreds::gitcreds_set()
```

Install `multideconv` from GitHub:

```{r, eval=FALSE}
# install.packages("pak")
pak::pkg_install("VeraPancaldiLab/multideconv")
```

## Tutorials

Step-by-step tutorials are available in the **Articles** section of the navigation bar:

- **Deconvolution with default methods** — run first-generation deconvolution and access built-in signatures
- **Cell type subgroup analysis** — derive cell-type subgroups and build a deconvolution dictionary
- **Single-cell deconvolution** — construct metacells and apply second-generation methods
- **Pseudo-bulk profiles and benchmarking** — create pseudo-bulk data, generate custom signatures, and benchmark performance
- **Machine learning workflows** — use subgroup features in cross-validated ML models

## Citation

If you use `multideconv` in a scientific publication, please cite:

> Hurtado, M., Essabbar, A., Khajavi, L., & Pancaldi, V. (2025). *multideconv – Integrative pipeline for cell type deconvolution from bulk RNAseq using first and second generation methods.* bioRxiv. <https://doi.org/10.1101/2025.04.29.651220>
